---
title: 'Reproducible Research: Peer Assessment 1'
author: "Enzo Alda"
date: "Sunday, January 18, 2015"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

### Load libraries

```{r}
library(psych);
library(sfsmisc);
```

Load data and augment columns 
```{r}
# Read the data (already unzipped after download)
data <- read.csv("activity.csv");
# Convert *date* field to proper Date
data$date <- as.Date(data$date);
# Add weekday column
data$weekday <- weekdays(data$date);
# Add the 'minute' column, showing the number of minutes from day start
data$minute <- as.integer(data$interval / 100) * 60 + (data$interval %% 100)
# Add a proper timestamp column
time.string <- paste(as.character(as.integer(data$interval / 100)), as.character(data$interval %% 100), sep=":");
data$time <- strptime(paste(data$date, time.string), format = "%Y-%m-%d %H:%M");
```

Check data types
```{r}
# Check data types
str(data);
# show summary
summary(data);
```

Compute mean and median for the total number of steps taken per day
```{r}
steps <- tapply(data$steps, data$date, sum);
mn <- mean(steps, na.rm = T)
md <- median(steps, na.rm = T)
```

Show mean and median
```{r}
mn
md
```

Plot histogram
```{r}
hist(steps, breaks=9, main="number of steps per day", xlab="number of steps", col="lightblue");
abline(v = mn, col="red", lwd=2);
```

Display average daily activity pattern
```{r}
n <- empty.dimnames(tapply(data$steps, data$interval, mean, na.rm = T));
t <- unique(data$interval);
steps <- data.frame(t=t, n=n);
plot(steps, type="l", xlab="time");
```

Locate maximum
```{r}
max.steps <- max(steps$n);
index <- match(max.steps, steps$n);
steps$t[index]
```
The interval 835 corresponds to the time interval 8:35-8:40 am

Calculate and report the total number of missing values in the dataset
```{r}
sum(is.na(data$interval))
```

Define function for imputing values
Our strategy would be to replace missing values (NA steps) with the mean number
of steps for the corresponding interval in the corresponding weekday.
```{r}
# function 'round.mean' returns the rounded value of the mean of a vector
round.mean <- function(v) {
    round(mean(v, na.rm=T));
};

# Overall estimates
estimates <- aggregate(steps ~ weekday + interval, data=data, FUN=round.mean);

# Prepare new data set
pdata <- data;

# Get number of rows
N <- dim(pdata)[1];

# Impute missing values
for (i in 1:N) {
    if (is.na(pdata$steps[i])) {
        pdata$steps[i] <- imputed.steps(pdata$weekday[i], pdata$interval[i]);
    }
}
```

```{r}
# hist mean median
steps <- tapply(pdata$steps, pdata$date, sum);
mn <- mean(steps, na.rm = T)
md <- median(steps, na.rm = T)
hist(steps, breaks=9, main="number of steps per day", xlab="number of steps", col="lightblue");
abline(v = mn, col="red", lwd=2);
mn
md

#
pdata$wday <- NULL;
pdata$wday[  pdata$weekday %in% c("Saturday", "Sunday") ] <- "weekend";
pdata$wday[!(pdata$weekday %in% c("Saturday", "Sunday"))] <- "workday";
pdata$wday <- as.factor(pdata$wday);

par(mfcol=c(1, 2));

plot.ts(pdata[pdata$wday == "workday",]$steps,
        main="Average Daily Steps\nWeekdays and Weekends compared", 
        ylab = "Average Steps", 
        xlab = "Time in Hours",
        axes = FALSE,
        col="red") 
axis(1,seq(0, 288, 24), label = seq(0,24, 2))
axis(2)
plot.ts(pdata[pdata$wday == "weekend",]$steps,
        axes=FALSE,
        xlab=NULL,
        ylab=NULL,
        col="blue") 
axis(1,seq(0, 288, 24), label = seq(0,24, 2))
axis(2)
legend("topright", c("Weekdays", "Weekends"), lty=c(1,1), col=c("red", "blue"))
```

